Making ice sheet models \\ scale properly

### 2023-01-21 talk/slides.tex
better slides

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glacier geometry and velocity evolve \emph{in contact with its climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating (\emph{ice shelf})
    \item[$\circ$] calving into the ocean
    \end{itemize}
\item and the \emph{topography} on which it sits
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\item<2> {\footnotesize next, a basic definition of ``ice sheet model''}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

    \item[$\circ$] \emph{bed elevation} $b(x,y)$
    \end{itemize}
\item two outputs:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x,y)$
    \item[$\circ$] \emph{ice velocity} $\bu(t,x,y,z)$
    \end{itemize}

@@

\begin{frame}{notation for the basic model}

\begin{itemize}
\item the \alert{data} $a(t,x,y)$, $b(x,y)$ \alert{is defined on a fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$

\begin{center}
\includegraphics<1>[width=0.55\textheight]{images/domain-data.png}
\includegraphics<2>[width=0.6\textheight]{images/domain-velocity.png}
\end{center}

\medskip
\item<2> the solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] $s=b$ where there is no ice
    \end{itemize}

\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \Omega$ on which the velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

\vspace{-2mm}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored in this talk}
    \end{itemize}
\item a \emph{key idea} is that conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}
    \end{itemize}
\item where $\Gamma_s(t)$ and $\Gamma_b(t)$, subsets of $\partial \Lambda(t)$, denote the surface and base of the ice
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item but ice sheet evolution is a \alert{free-boundary} problem for conserved quantities in a (thin) fluid layer
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the fixed map-plane domain $\Omega\subset \RR^2$, glacier \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

\bigskip
\item<2> {\footnotesize for more on this free-boundary perspective see Bueler (2021)}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model (strong form)}

@@

@@

\begin{itemize}
\item a thickness-transport form exposes evolution type
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<2-> \emph{question:} is it really an advection?
\item<2->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
\end{itemize}
\end{frame}

\begin{frame}{shallow ice sheet models}

\begin{itemize}
\item the ``basic model'', i.e.~NCP coupled to Stokes, is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] for now, it is too expensive at scale
    \item[$\circ$] goal: \alert{make it fast enough to use!}
    \end{itemize}

\smallskip
\item<2-> current-technology models use various shallow simplifications
\item<2-> simplest is a lubrication approximation, the \alert{shallow ice approximation}, in which $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator

\smallskip
\item<3-> the diffusion form of mass continuity ($\frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) = a$) is only literally valid in the lubrication approximation, but \emph{the fact that ice flows downhill has time-stepping stability consequences}
\end{itemize}
\end{frame}

@@

{\small
\begin{align*}
s - b &\ge 0 \\
\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
\end{align*}
where
\begin{align*}
\Phi(s) &= - \bu|_s \cdot \bn_s \\
        &= - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
\end{align*}

\begin{itemize}
\item $\nabla$ only acts in horizontal variables
\item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation
\item $\Phi$ is a \alert{doubly-nonlinear differential operator}, simultaneously of porous medium and $\pp$-Laplacian type
\end{itemize}
}
\end{frame}

@@

{\footnotesize
\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}

\begin{itemize}
\item recall some traditional numerical analysis \dots
\end{itemize}
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item we expect \alert{explicit} time stepping for advections
\item for example, forward Euler in 1D gives
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bQ_{j+1/2}^\ell - \bQ_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] where $\bQ = \bU H$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<2>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler in 1D gives
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

@@

\begin{frame}{time-stepping in ice sheet models}

\begin{itemize}
\item current-technology models mostly ``believe'' that the problem is an advection and use explicit time stepping
\item FIXME but this has performance consequences
\item FIXME the problem is a DAE!
\item FIXME implicit alternative requires solving step equations: simultaneous NCP for SKE and Stokes
\end{itemize}
\end{frame}

\section{on Stokes solver scaling}

\begin{frame}{regarding PDE solver complexity}

FIXME
\end{frame}

\section{a simplified performance analysis}

\begin{frame}{complexity of ice sheet models}

\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
\onslide<1->{
\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
}
\onslide<1->{
explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
}
\onslide<2->{
explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
}
\onslide<3->{
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
}
\onslide<3->{
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
}
\end{tabular}

@@

\section{scalable implicit steps by multigrid?}

\begin{frame}{obstacle problem solver scaling}

FIXME
\end{frame}

>>>>

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt when floating
    \item[$\circ$] calving into the ocean
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item at least two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$\hspace{-7mm} a(t,x,y) = \begin{pmatrix}
\text{precipitation minus} \\
\text{melt \& runoff}
\end{pmatrix}$$

@@

    \item[$\circ$] \emph{bed elevation} $b(x,y)$
    \end{itemize}
\item at least two outputs:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x,y)$
    \item[$\circ$] \emph{ice velocity} $\bu(t,x,y,z)$
    \end{itemize}

@@

\begin{frame}{basic ice sheet model: notation}

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item the \alert{data} $a(t,x,y)$, $b(x,y)$  is defined on a fixed domain:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2> the solution \alert{surface elevation} $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] $s=b$ where there is no ice
    \end{itemize}

\medskip
\item<2> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$ on which the solution \alert{velocity} $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

\vspace{-2mm}
\end{itemize}
\end{column}
\begin{column}{0.4\textwidth}
\includegraphics<1>[width=\textwidth]{images/domain-data.png}

\includegraphics<2>[width=\textwidth]{images/domain-velocity.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item ice sheet evolution should conserve physical quantities:
    \begin{itemize}
    \item[$\circ$] mass
    \item[$\circ$] momentum
    \item[$\circ$] \st{energy} \hfill $\leftarrow$ \emph{ignored for simplicity in this talk}
    \end{itemize}
\item a key idea is that conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

    \item[$\circ$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
    \end{itemize}

\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet evolution is a \alert{free-boundary} problem for conserved mass and momentum quantities
\item specifically, the surface kinematic equation
  $$\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s = a$$
applies \emph{only} on the ice upper surface $\Gamma_s(t)$
\item in the remainder of the (fixed, map-plane) domain $\Omega\subset \RR^2$, \alert{complementarity} holds:
  $$s=b \quad \text{and} \quad a \le 0$$

\bigskip
\item<2> {\footnotesize for more on this perspective see Bueler (2021)}
\end{itemize}
\end{frame}

\begin{frame}{basic ice sheet model: strong form}

@@

\begin{frame}{basic ice sheet model: its a DAE system anyway}

\begin{itemize}
\item for clarity and simplicity, declutter by forgetting complementarity and all the other boundary conditions, too \dots
\item get surface kinematic equation coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 && \text{on } \Omega \subset \RR^2 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''}
\end{align*}
\item observe that $s(t,x,y)$ and $\bu(t,x,y,z)$ both evolve in time
\item \emph{however}, only the first of the 5 equations has a time derivative
\item the problem is a \alert{differential algebraic equation} system, e.g.
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item[$\circ$] \dots actually in infinite dimensions (\emph{PDAE}?), and subject to complementarity (\emph{free boundary problem for PDAE})
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{important context (current research)}

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this ``NCP coupled to Stokes'' glacier evolution model
    \begin{itemize}
    \item[$\circ$] however, I think most researchers would agree this is the intended model!
    \item[$\circ$] well-posedness of shallow approximations has been considered\footnote{e.g.~Jouvet \& Bueler (2012)}
    \end{itemize}

\medskip
\item<2> numerical modelers tend think of the Stokes problem separately from surface evolution (\emph{time-split} or \emph{explicit time-stepping}),
\item<2> and glacier evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item thickness-transport form helps expose evolution character
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note: $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item<2-> the \alert{mass continuity equation} follows from the surface kinematical equation and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<3-> \emph{question:} is it really an advection?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective form:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion form:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item both forms are highly-nonlinear: \, $\bU=\bU(H,\grad s), \, D = D(H,\grad s)$
\item the diffusion form is only \emph{literally} valid in the lubrication approximation (more below)
\item the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences} regardless
\end{itemize}
\end{frame}

\begin{frame}{numerical shallow ice sheet models}

\begin{itemize}
\item the ``basic ice sheet model'' (NCP coupled to Stokes) is \emph{not yet} in common use for ice sheet simulations
    \begin{itemize}
    \item[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
    \item[$\circ$] goal: \alert{make it fast enough to use!}
    \end{itemize}

\smallskip
\item<2-> current-technology models use various shallow simplifications
    \begin{itemize}
    \item<2->[$\circ$] usual justification for falling back to shallow approximations: ``Stokes is too expensive at scale''
    \end{itemize}
\item<2-> the simplest shallow approximation is a lubrication approximation, the \alert{shallow ice approximation} (SIA)
\item<2-> in the SIA, $- \bu|_s \cdot \bn_s$ resolves to a nonlinear differential operator

\smallskip
\end{itemize}
\end{frame}

@@

{\small
\begin{align*}
s - b &\ge 0 \\
\frac{\partial s}{\partial t} + \Phi(s) - a &\ge 0 \\
(s - b) \left(\frac{\partial s}{\partial t} + \Phi(s) - a\right) &= 0
\end{align*}
\begin{equation*}
\Phi(s) = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)
\end{equation*}
}

\begin{itemize}
\item constants $\pp = \nn+1$ and $\gamma > 0$ relate to ice deformation

\medskip
\item<2> thus: $\Phi(s) = - \bu|_s \cdot \bn_s$ resolves to a \alert{doubly-nonlinear} differential operator, of porous medium and $\pp$-Laplacian type simultaneously
\item<2> Jouvet \& Bueler (2012): existence for this problem
\end{itemize}
\end{frame}

\AtBeginSection[]
{
  \begin{frame}{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

@@

{\footnotesize
\begin{align*}
\text{advective ice sheet model schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion ice sheet model schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}

\bigskip
\begin{itemize}
\item recall: traditional numerical analysis applied to these equations
\end{itemize}

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{\item[]}
\only<2>{
\item we expect \alert{explicit} time stepping for advections
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{\Delta x} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] maximum time step from CFL using $\max |\bU|$
    \end{itemize}
}
\only<3>{
\item we expect \alert{implicit} time stepping for diffusions
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (H_{j+1}^{\ell+1} + H_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (H_{j}^{\ell+1} + H_{j-1}^{\ell+1})}{2 \Delta x} = a_j^\ell$$

@@

\begin{frame}{time-stepping in current ice sheet models}

\begin{itemize}
\item current-technology models mostly ``believe'' the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] an embarrassing situation \dots the problem is a DAE!
    \item[$\circ$] performance consequences: stability is determined by the coupled solution velocity $\bU$ not a pre-determined velocity
    \end{itemize}

\medskip
\item<2> an alternative should be obvious to any numerical analyst \dots
\item<2> \alert{implicit time-stepping} is expected for DAE problems
\item<2> implicit steps imply a sequence of problems (``equations'') to solve at each time step
\item<2> here: implicit $\implies$ NCP coupled to Stokes at each time step
\end{itemize}
\end{frame}

\section{Stokes solver scaling}

\begin{frame}{PDE solver complexity}

FIXME explain in terms of 2D Poisson
\end{frame}

\begin{frame}{a scalable Stokes solver for ice sheets}

FIXME Isaac et al.~2015
\end{frame}

\begin{frame}{a scalable Stokes solver for ice sheets 2}

FIXME table from Isaac, and estimate power $\alpha$
\end{frame}

\section{simplified performance analysis}

\begin{frame}{analyzing ice sheet model performance}

@@

\begin{frame}{ice sheet model performance: table}

\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$
\end{tabular}

FIXME table from Bueler (2022)
\end{frame}

\begin{frame}{complexity for implicit solves}

FIXME $\beta$, $\gamma$
\end{frame}

\begin{frame}{ice sheet model performance: table cont.}

\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
\end{tabular}

FIXME table from Bueler (2022)
\end{frame}

\section{scalable approaches: candidates}

\begin{frame}{semi-coupled time-stepping}

FIXME L{\"o}fgren et al.~2022
\end{frame}

\begin{frame}{machine learning}

FIXME Jouvet et al.~2021
\end{frame}

\begin{frame}{scalable obstacle problem solvers}

FIXME chapter 9 from my book
\end{frame}

\begin{frame}{scalable obstacle problem solvers}

FIXME multilevel multigrid approach
\end{frame}

\section{conclusion}
